\titletext{}

### 2020-10-21 manuscript/regional_befr_manuscript.tex
Updated tex

\title{Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}
\author{}
\date{}

@@

\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}

@@

\textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear. It is posited that disturbance moderates the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function. We explore the relationship between tree species diversity and biomass across southern African woodlands, accounting for variation in stem density, resource availability, disturbance, and vegetation type to build general understanding of the biodiversity-ecosystem function relationship in this understudied ecological context.

\textbf{Location:} Southern African savannas and woodlands

\textbf{Time period:} 2010-2019

\textbf{Major taxa studied:} Trees

\textbf{Methods:} We used a network of \nplots{} woodland tree plots located across southern Africa. We used Structural Equation Modelling to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.

\textbf{Results:} We found positive effects of tree species diversity on aboveground biomass, across vegetation types, observed mainly indirectly via the increasing effect of woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}. Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.

\textbf{Main conclusions:} The study underlines the close association between tree diversity and ecosystem structure and function of highly disturbed southern African savannas and woodlands. Our results demonstrate the importance of accounting for environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.

@@

% BEFR has had previous global research
In order to understand the effects of global biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}. The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016, Hooper2012, Cardinale2009}. Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}, which follow from early small-scale experimental studies conducted predominantly in artificial grassland mesocosms \citep{Tilman1994, Tilman2014}. Despite these concerted efforts, we continue to lack a nuanced, ecosystem-agnostic, understanding of the complex interactions between biodiversity, environment, and ecosystem function.

@@

% In this study :::
In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots spanning broad environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees as our metric of ecosystem function, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results. We used Structural Equation Modelling (SEM) with path analysis to simultaneously account for environmental and biotic factors, which may have interacting effects on ecosystem structure and therefore biomass. Initially, we posited three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with plot-level stem density (number of stems ha\textsuperscript{-1}), due to an increased importance of niche complementarity as stem density and therefore competition increases. In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Re-cap
In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots. Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}). Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}). We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} below which the effect of species diversity on AGB remained at a low baseline level. The strongest effect on AGB was that of stem density. When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.

@@

In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB). Additionally, we found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees. We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar ecosystem processes occur across the region to determine ecosystem function. In contrast to previous studies, we found, at the scale of our study region, that the direct effects of water availability and soil fertility on woody biomass were small, with most of their effect being indirectly through their effects on tree species and structural diversity. This strongly suggests that data on tree species diversity be included into models predicting ecosystem functionality in this region. We also advocate for explicit inclusion of environmental covariates in regional scale models of biodiversity and ecosystem function, generally.

The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in favour of disturbance effects. On a practical note, our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling. This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.

>>>>

\newcommand{\titletext}{Stem density drives biodiversity-ecosystem function across disturbed southern African woodlands}

\title{\titletext{}}
\author{}
\date{}

@@

\LARGE{Title: \titletext{}}

@@

\begin{itemize}
	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but whether BEFRs should be expected in disturbance-driven systems is less clear. Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs. We explore the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.}
	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity. The observed BEFR was highly dependent on stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}. We also found that resource availability mainly affects biomass indirectly, via increasing species diversity.}
	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed southern African savannas and woodlands. We suggest that tree diversity is an under-appreciated determinant of savanna and woodland structure and function.}
\end{itemize}

@@

% BEFR has had previous global research
In order to understand the interacting effects of global environmental and biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}. The strength and direction of the Biodiversity-Ecosystem Function relationship (BEFR) varies depending on the ecosystem studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016, Hooper2012, Cardinale2009}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}, which follow from early small-scale experimental studies conducted predominantly in temperate grasslands \citep{Tilman1994, Tilman2014}. Despite these concerted efforts, we continue to lack a nuanced, ecosystem-agnostic, understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.

@@

% In this study :::
In this study, we make the first known estimation of the biodiversity-ecosystem function relationship (BEFR) across a sub-continental area of disturbance-driven ecosystems, namely southern African savannas and woodlands. We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to seeking a general model of the biodiversity-ecosystem function relationship in wooded ecosystems. We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), due to an increased importance of niche complementarity as stem density and therefore competition increases. In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass. Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e. physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.

@@

% Re-cap
In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM). We found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}). Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}). We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} below which the effect of species diversity on AGB remained at a low baseline level. The strongest direct effect on AGB was that of stem density. When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.

@@

In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB). 
Our findings contribute to our understanding of a universal biodiversity-ecosystem function relationship, one which is moderated in a predictable manner by environmental covariates and their interaction with biodiversity and ecosystem structure.
We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar  processes operate across the region to determine ecosystem function. 
We advocate for explicit inclusion of environmental covariates in regional scale models of biodiversity and ecosystem function. We assert that this is necessary to develop our understanding of the biodiversity-ecosystem function relationship in real-world ecosystems, to progress from experimental mesocosms.
We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key vector through which tree species diversity determines ecosystem function in savannas, woodlands and forests, i.e. all wooded ecosystems.
The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in previous studies despite its intuitive logic as a determinant of niche complementarity effects in wooded ecosystems.
Our study shows that biodiversity change through extensive human actions in this region will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling. This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
